\documentclass{tufte-handout}
\usepackage{graphicx}
\usepackage{url}
\usepackage{enumerate}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{enumitem}
\usepackage[iso,american]{isodate}
\usepackage{fancyhdr}
\fancypagestyle{firstpage}{
  \rhead{Demos and Synthesis \linebreak
  \textit{Version: \today}}
}

\title{Module 3 and QEA Synthesis and Reflection}
\author{Quantitative Engineering Analysis}
\date{}

\begin{document}

\maketitle
\thispagestyle{firstpage}

\begin{abstract}
\end{abstract}

\section{Schedule}
\begin{itemize}
\item 8:45-9:00: Coffee and Donuts
\item 9:00-9:30: Live Demo and Debrief on Module 3
\item 9:30-10:15: QEA I Synthesis
\item 10:15-11:00 Reflection/Retrospective/Feedback Survey
\end{itemize}


\section{Live Demo and Debrief on Module 3 [30 minutes]}

\subsection{Live Demo or Video Screening}  
If you would like to Live Demo your running of the Gauntlet, now is the time!!!  Also, if you have a video you'd like to share, we'd love to see it!

\subsection{Debrief on Module 3}
\begin{itemize}
\item \newthought{Beginning with the Gauntlet} and looking back through the deliverables from Module 3, create a map of each of the challenges of Module 3 (BoD, Flatland, the Gauntlet).  Identify key concepts, tools, processes, skills, etc. from each challenge and how these relate to each other.  Any items which link to more than one Challenge should be used as linkers to connect the individual challenge maps into a single web of connections.  How many connections between the challenges can you identify?  Document your map.
\item \newthought{As you work through the map} use this opportunity to organize your files from Module 3.  Make sure you have copies of all the assignment documents as well as your own work in one place that is well organized so you can find things.  WE WILL BE USING A LOT OF THESE CONCEPTS AGAIN IN THE FALL!!!
\item  \newthought{Now, compare maps} with the other teams working on the same wall as you are.  Working with these other teams, create one map for Module 3 which includes all the concepts, tools, processes, skills and connections from all the groups.  This map should use about 1/3 of board space available to you!  Consider color coding your map based on concepts, tools, processes, "soft skills," etc. Don't forget about context and consequences.
\end{itemize}

\section{QEA I Debrief [45 minutes]}
\begin{itemize}
\item \newthought{Dividing the group at your board into two teams} and looking back through the deliverables from the previous modules, create a map for each of Module 1 and Module 2 in the remaining 2/3 of your board space.  Identify all key concepts, tools, processes, skills, etc. and how these relate to one another.
\item \newthought{As you work through the map} use this opportunity to organize your files from Modules 1 and 2.  Make sure you have copies of all the assignment documents as well as your own work in one place that is well organized so you can find things.  WE WILL BE USING A LOT OF THESE CONCEPTS AGAIN IN THE FALL!!!
\item \newthought{Swap groups} working on Modules 1 and 2.  Can you find ideas or connections that the previous group missed?
\item \newthought{Again working as a team with the whole group sharing your board} Make connections between all three modules.  How many connections can you find?  Connections can be concepts, tools, processes, skills, etc. What are the points of largest connection across the whole class with respect to concepts, tools, processes, skills, etc.?  (In other words, what is the most connected concept?  The most frequently encountered tool, etc.?)
\item \newthought{Document your group's map of the class} and make sure each teammember has access to that document.  Take time to browse both rooms and take a look at other group's maps.  Did they find connections you didn't see?
\end{itemize}


\section{Reflection and Feedback [45 minutes]}

\newthought{Now it is time} for you personally to look back on the semester.

\begin{itemize}
\item On Canvas, there is a survey for you to fill out.  We ask you to take the time to reflect on your own personal development during this semester, and what you would like to work on improving during the coming semester.  Remember, we are only partway through the class, not at the end!  We also would very much like your feedback on the class and anything you would like us to work on as we head into the second semester. (If you would like to leave anonymous feedback, the anonymous feedback form is still \href{https://docs.google.com/forms/d/e/1FAIpQLSefl9-C9qsvMEPz8al1eZWOV5sD7Gqfmmyzy-zn87bWm1-jvQ/viewform}{here}.)  Please finish and submit the end-of-semester survey before 11. 
\item Please also fill out the course evaluation through the Olin portal if you have not done so already.
\end{itemize}

\newthought{Have a great summer and we'll see you in the fall!}


\end{document} 